Bayesian Hierarchical Modeling: An Introduction and Reassessment

### 2023-04-06 manuscript.Rmd
Updates revision

Before diving into hierarchical modeling, let us consider two alternative approaches: Analyzing aggregated data and analyzing the data separately for each participant. Analyses of aggregated data such as the ANOVA from the scenario above examine a general effect that is assumed to be consistent across subjects. For instance, when investigating whether there is an effect of digit on RT, the mean RT for each participant is computed for each digit-condition, and a repeated-measures ANOVA is conducted on these aggregated scores. Analyses of aggregated data cannot investigate whether a trend is consistent across participants. Therefore, when this option is applied and individual variability is ignored, the general trend can be an inaccurate representation of the true general trend [Rouder, Morey, & Pratte, 2013; Haaf & Rouder, 2017; van Doorn, Aust, Haaf, Stefan, and Wagenmakers, -@van2021bayes]. This is shown in Panel A of Figure \@ref(fig:figure1) where some of the individual trends (i.e., slopes of the grey lines) differ from the general trend (i.e., slope of the blue line). If every participant takes part in each condition, the person variability affects all conditions, thus inducing a correlation across conditions. For example, participants in a digit classification task repeatedly respond to target digits inducing a correlation between the responses to the target digits, such as between target digit 3 and target digit 4. Therefore, the observations are not independent as we assume if we perform analysis on aggregated data. As a result, the type I error rate of a test of the effect increases: there is a higher probability of concluding that there is an effect if there is no effect (Rouder & Lu, 2005).

@@

### Model Fitting 
The model is fitted using the packages `rstan` and `brms`. For `rstan`, four chains are run with 4000 iterations per chain, of which 1000 serve as warm-up. This results in a total of 12000 samples from the joint posterior distribution. For `brms`, four chains are run with 6000 iterations per chain, of which 1000 serve as warm-up. This results in a total of 20000 draws from the joint posterior distribution.

@@

For the symbolic distance effect, the estimated number of effective samples is provided in panel D of Figure \@ref(fig:traceplots3p). For the package `rstan` the number is generally high, much higher than the actual number of samples depicted by the green dashed line. The `brms` package shows much more variation. The lowest effective sample sizes can be found for the intercept parameters. Table 1 also displays the number of effective samples for the general effects parameters in the `rstan` model. If the number of effective samples is too low, `brms` and `rstan` will return a warning. Here, this was not case, and all other diagnostics were good, so we continue with the interpretation of the estimates.

@@

For the digit effects, positive parameter estimates indicate that the response is slower for the non-baseline digits, that is, digits that are closer to 5. The parameters $\mu_{\delta_{6}}$ and $\mu_{\delta_{4}}$ are distributed around higher values and their credible intervals do not contain zero. Note that this result is consistent with the symbolic distance hypothesis that the digits 4 and 6 have the highest response times because they are closest to the 5, the comparison value. The 95% credible intervals of $\mu_{\delta_{7}}$ and $\mu_{\delta_{3}}$ contain zero. \Revision{Therefore, it is plausible that the response times do not differ substantially from the baseline condition.- is this one necessary to adjust?} The posterior distribution of the side effect $\mu_{\beta}$ is centered around zero with a posterior mean of `r printnum(fitrstanhiertable3["mu2", "Mean"], digits = 3)` and a 95% credible interval ranging from `r printnum(fitrstanhiertable3["mu2", "Lower Bound"], digits = 3)` to `r printnum(fitrstanhiertable3["mu2", "Upper Bound"], digits = 3)`. \Revision{This indicates a small to non-existent side effect. -change}

@@

The largest variation across participants can be seen for the intercept parameter. This means that participants vary considerably in their baseline response times. There is also considerable individual variation in the digit effects ($\delta$'s). However, most of the intervals contain zero which implies that the effects might be small or non-existent. The variability is most pronounced for the effect parameters of the digits close to 5. For these digits, there are also more intervals that do not contain zero, indicating that an effect might be present on the subject level. The posterior mean of the individual side effect ($\beta_{i}$) varies between -0.05 and 0.05. Again, all but four 95% credible intervals include zero indicating uncertainty about the existence of an effect for anyone. The credible intervals of the side effect and digit effects $\delta_{7}$ and $\delta_{3}$ indicate that different subjects may show opposite effects [@haaf2019some].

@@

```{r modelestimatesfigure3, echo=FALSE, fig.cap="Model estimates for digit effect parameters in \\textit{rstan}. The points represent the mean parameter estimates for each individual. The violin plots on the right side show the variance of the individual parameter estimates. The size of effects is not very constistent within a person.", fig.align = "left", fig.height = 11, fig.width = 1, out.width="75%"}
knitr::include_graphics("Images/Fig11_modelestimatefigure2_onlyrstan_v2.png")
```

@@

\begin{equation}
\begin{aligned}
&e^{(\mu_{\gamma} + x_j \mu_{\beta} + w_j \mu_{\delta_{3}})} - e^{(\mu_{\gamma} +  x_j \mu_{\beta})} \\
&= e^{(-.56 + .5 \times (-.01) + 1 \times 0.03)} - e^{(-.56 + .5 \times (-.01))} \\ 
&= 0.017 
\end{aligned}
\end{equation}
This value corresponds to the median estimate of $\mu_{\delta_{3}}$ from the normal model. Calculating the mean estimate of $\mu_{\delta_{3}}$ would be similar, but more complicated as the mean of a log-normal distribution also depends on its variance.

One property of these parameter estimates we can immediately interpret is their sign (i.e., positive, negative or zero). A positive estimate indicates that the RT increases, whereas a negative effect indicates a decrease. When inspecting the general digit effect parameters, they all appear to positively influence the RT. This is in line with our expectations since the baseline RT in the model is set equal to the RT for the digits furthest away from 5 and the symbolic distance effect postulates that RT should increase as the digits become closer to five. The side effect is close to zero, with a posterior mean of `r printnum(fitrstanhiertable3log["mu2", "Mean"], digits = 3)` and a 95% credible interval from `r printnum(fitrstanhiertable3log["mu2", "Lower Bound"], digits = 3)` to `r printnum(fitrstanhiertable3log["mu2", "Upper Bound"], digits = 3)`. Thus, the general effect of side seems small or negligible.

@@

We assess the stability of the obtained Bayes factor by repeating the estimating process ten times with `rstan` and `brms`, respectively. The results are shown in Figure \@ref(fig:bsbfestimatesbrmsrstan). The figure shows the evidence in favor of the null, digit and side model against the full model. All Bayes factors are larger than one. Therefore, it can be concluded that the data are more likely under less complex models. There is almost no variation in the computed Bayes factors per model, indicating that the Bayes factor estimate is stable. However, the Bayes factor estimates in `rstan` generally show stronger evidence than Bayes factor estimates based on `brms`. It is important to note that this difference is not due to the bridge sampling function, but a result of a difference in parametrization. However, for both packages, the conclusion is the same: the data are least likely to have occurred under the full model. In case the estimates are unstable, it is recommended to further increase the number of posterior samples.

@@

To obtain the posterior model probability for each model in every scenario, we first compute the Bayes factor of every model against the null model. This includes the trivial Bayes factor of the null model against the null model, resulting in ten Bayes factors in total. Next, we calculate the sum of all these Bayes factors and divide each Bayes factor by this sum. This results in the posterior model probabilities shown in Table \@ref(tab:postmodprob).^[Note that it would be possible to calculate the posterior model probabilities directly from the marginal likelihoods. However, marginal likelihoods are often not computed explicitly in the model fitting process (e.g., when using the bayes_factor function in the `brms` package), so we believe that readers might find it easier to compute posterior model probabilities from Bayes factors.] All posterior model probabilities from the table and the posterior probability of the null model (see table note) sum to one.

@@

The table shows that the digit model without individual digit effects from scenario 3 has the highest posterior model probability, followed by the full model without individual digit and side effects (scenario 3). This means that the data are most likely under these models than the other eight models included in the comparison. In addition, the models with smaller prior expectations for individual differences (scenario 2), have higher posterior model probabilities than the models in scenario 1. Thus, decreasing our expectations of the size of the individual digit and side effects, or completely removing them, results in higher posterior model probabilities. Together with Figure \@ref(fig:randomparfigure), showing the small differences in individual effects (also illustrated by Haaf et al., 2019), this result supports the notion that side and digit effects are small and there is a lack in individual variability.

@@

## Potential Technical Issues

@@

Another technical issue may arise with the estimation of Bayes factors. Here, we highlighted two approaches, the Savage-Dickey density ratio [@wagenmakers2010bayesian] and bridge sampling [@gronau2017tutorial]. We already highlighted the drawbacks of the SD density ratio. Yet, bridge sampling may also fail in some cases. Failures of the bridge sampling algorithm are identified by unstable Bayes factors across repeated sampling runs. There are improvements to the algorithm that are specifically developed for hierarchical models [@gronau2019simple]. Yet, it remains crucial to assess the stability of Bayes factor estimates.

>>>>

Before diving into hierarchical modeling, let us consider two alternative approaches: Analyzing aggregated data and analyzing the data separately for each participant. Analyses of aggregated data such as the ANOVA from the scenario above examine a general effect that is assumed to be consistent across subjects. For instance, when investigating whether there is an effect of digit on RT, the mean RT for each participant is computed for each digit-condition, and a repeated-measures ANOVA is conducted on these aggregated scores. Analyses of aggregated data cannot investigate whether a trend is consistent across participants. Therefore, when this option is applied and individual variability is ignored, the general trend can be an inaccurate representation of the true general trend [Rouder, Morey, & Pratte, 2013; Haaf & Rouder, 2017; van Doorn, Aust, Haaf, Stefan, and Wagenmakers, -@van2021bayes]. This is shown in Panel A of Figure \@ref(fig:figure1) where some of the individual trends (i.e., slopes of the grey lines) differ from the general trend (i.e., slope of the blue line). If every participant takes part in each condition, the person variability affects all conditions, thus inducing a correlation across conditions. For example, participants in a digit classification task repeatedly respond to target digits inducing a correlation between the responses to the target digits, such as between target digit 3 and target digit 4. \Revision{Therefore, analyses on aggregated data do not appropriately account for the interdependence between observations [@van2021bayes; @van2023bayes].} As a result, the type I error rate of a test of the effect increases: there is a higher probability of concluding that there is an effect if there is no effect (Rouder & Lu, 2005).

@@

### Model Fitting 
The model is fitted using the packages `rstan` and `brms`. \Revision{We used different settings for `rstan` and `brms` due to the difference between the models for these packages. Although we tried to make the models as similar as possible, the models are not identical. This is partly due to difference in prior specification (i.e., specifying prior on the variances versus standard deviations).} For `rstan`, four chains are run with 4000 iterations per chain, of which 1000 serve as warm-up. This results in a total of 12000 samples from the joint posterior distribution. For `brms`, four chains are run with 6000 iterations per chain, of which 1000 serve as warm-up. This results in a total of 20000 draws from the joint posterior distribution.

@@

For the symbolic distance effect, the estimated number of effective samples is provided in panel D of Figure \@ref(fig:traceplots3p). For the package `rstan` the number is generally high, much higher than the actual number of samples depicted by the green dashed line. The `brms` package shows much more variation. The lowest effective sample sizes can be found for the intercept parameters. Table 1 also displays the number of effective samples for the general effects parameters in the `rstan` model. If the number of effective samples is too low, `brms` and `rstan` will return a warning. Here, this was not case, and all other diagnostics were good, so we continue with the interpretation of the estimates. \Revision{The difference between the model diagnostics of `rstan` and `brms` can be partly explained by the difference in model parametrization, such as the difference in the prior specification (i.e., specifying priors on variances versus priors on standard deviations). Although we tried to make the models as similar as possible, they are not identical.}

@@

For the digit effects, positive parameter estimates indicate that the response is slower for the non-baseline digits, that is, digits that are closer to 5. The parameters $\mu_{\delta_{6}}$ and $\mu_{\delta_{4}}$ are distributed around higher values. Note that this result is consistent with the symbolic distance hypothesis that the digits 4 and 6 have the highest response times because they are closest to the 5, the comparison value. The posterior distribution of the side effect $\mu_{\beta}$ is centered around zero with a posterior mean of `r printnum(fitrstanhiertable3["mu2", "Mean"], digits = 3)` and a 95% credible interval ranging from `r printnum(fitrstanhiertable3["mu2", "Lower Bound"], digits = 3)` to `r printnum(fitrstanhiertable3["mu2", "Upper Bound"], digits = 3)`. \Revision{Note that the credible interval overlaps with zero. Commonly, this is misinterpreted as an absence of an effect. However, in combination with the mean estimate, we can conclude that the effect is small.}

@@

The largest variation across participants can be seen for the intercept parameter. This means that participants vary considerably in their baseline response times. There is also considerable individual variation in the digit effects ($\delta$'s). \Revision{However, most of the intervals contain zero which implies that the effects might be small or non-existent.} The variability is most pronounced for the effect parameters of the digits close to 5. For these digits, there are also more intervals that do not contain zero, indicating that an effect might be present on the subject level. The posterior mean of the individual side effect ($\beta_{i}$) varies between -0.05 and 0.05. Again, all but four 95% credible intervals include zero indicating uncertainty about the existence of an effect for anyone. The credible intervals of the side effect and digit effects $\delta_{7}$ and $\delta_{3}$ indicate that different subjects may show opposite effects [@haaf2019some].

@@

```{r modelestimatesfigure3, echo=FALSE, fig.cap="Model estimates for digit effect parameters in \\textit{rstan}. The points represent the mean parameter estimates for each individual. The violin plots on the right side show the variance of the individual parameter estimates. ", fig.align = "left", fig.height = 11, fig.width = 1, out.width="75%"}
knitr::include_graphics("Images/Fig11_modelestimatefigure2_onlyrstan_v2.png")
```

@@

\begin{equation}
\begin{aligned}
&e^{(\mu_{\gamma} + x_j \mu_{\beta} + w_j \mu_{\delta_{3}})} - e^{(\mu_{\gamma} +  x_j \mu_{\beta})} \\
&= e^{(-.56 + .5 \times (-.01) + 1 \times 0.03)} - e^{(-.56 + .5 \times (-.01))} \\ 
&= 0.017 
\end{aligned}
\end{equation}
This value corresponds to the median estimate of $\mu_{\delta_{3}}$ from the normal model. Calculating the mean estimate of $\mu_{\delta_{3}}$ would be similar, but more complicated as the mean of a log-normal distribution also depends on its variance. \Revision{This approach can also be used iteratively. This means that the approach above is applied to every iteration in the chain. This results in a posterior distribution of the effect quantifying the posterior uncertainty.}

One property of these parameter estimates we can immediately interpret is their sign (i.e., positive, negative or zero). A positive estimate indicates that the RT increases, whereas a negative effect indicates a decrease. When inspecting the general digit effect parameters, they all appear to positively influence the RT. This is in line with our expectations since the baseline RT in the model is set equal to the RT for the digits furthest away from 5 and the symbolic distance effect postulates that RT should increase as the digits become closer to five. The side effect is close to zero, with a posterior mean of `r printnum(fitrstanhiertable3log["mu2", "Mean"], digits = 3)` and a 95% credible interval from `r printnum(fitrstanhiertable3log["mu2", "Lower Bound"], digits = 3)` to `r printnum(fitrstanhiertable3log["mu2", "Upper Bound"], digits = 3)`. \Revision{This indicates that the general effect of side is small.}

@@

We assess the stability of the obtained Bayes factor by repeating the estimating process ten times with `rstan` and `brms`, respectively. The results are shown in Figure \@ref(fig:bsbfestimatesbrmsrstan). The figure shows the evidence in favor of the null, digit and side model against the full model. All Bayes factors are larger than one. Therefore, it can be concluded that the data are more likely under less complex models. There is almost no variation in the computed Bayes factors per model, indicating that the Bayes factor estimate is stable. However, the Bayes factor estimates in `rstan` generally show stronger evidence than Bayes factor estimates based on `brms`. It is important to note that this difference is not due to the bridge sampling function, but a result of a difference in parametrization. \Revision{The models for `rstan` and `brms` are similar but not identical, partly due to the difference in prior specification (i.e., variances versus standard deviations).} However, for both packages, the conclusion is the same: the data are least likely to have occurred under the full model. In case the estimates are unstable, it is recommended to further increase the number of posterior samples.

@@

To obtain the posterior model probability for each model in every scenario, we first compute the Bayes factor of every model against the null model. This includes the trivial Bayes factor of the null model against the null model. \Revision{This setup results in 10 models, and, therefore, 10 Bayes factors. We assume that all models are equally likely, resulting in a prior model probability for each model of 1/10.} Next, we calculate the sum of all these Bayes factors and divide each Bayes factor by this sum. This results in the posterior model probabilities shown in Table \@ref(tab:postmodprob).^[Note that it would be possible to calculate the posterior model probabilities directly from the marginal likelihoods. However, marginal likelihoods are often not computed explicitly in the model fitting process (e.g., when using the bayes_factor function in the `brms` package), so we believe that readers might find it easier to compute posterior model probabilities from Bayes factors.] All posterior model probabilities from the table and the posterior probability of the null model (see table note) sum to one.

@@

The table shows that the digit model without individual digit effects from scenario 3 has the highest posterior model probability, followed by the full model without individual digit and side effects (scenario 3). This means that the data are most likely under these models than the other eight models included in the comparison. In addition, the models with smaller prior expectations for individual differences (scenario 2), have higher posterior model probabilities than the models in scenario 1. \Revision{Thus, decreasing the variability of the size of the individual digit and side effects (scenario 2), or completely removing the variance (scenario 3), results in higher posterior model probabilities.} Together with Figure \@ref(fig:randomparfigure), showing the small differences in individual effects (also illustrated by Haaf et al., 2019), this result supports the notion that side and digit effects are small and there is a lack in individual variability.

@@

\Revision{## Technical Possibilities are Endless}

@@

Another technical issue may arise with the estimation of Bayes factors. Here, we highlighted two approaches, the Savage-Dickey density ratio [@wagenmakers2010bayesian] and bridge sampling [@gronau2017tutorial]. We already highlighted the drawbacks of the SD density ratio. \Revision{Bayes factor estimates can also be biased. However, bridge sampling has shown to be relatively unbiased [@sarafoglou2021evaluating].} Yet, bridge sampling may also fail in some cases. Failures of the bridge sampling algorithm are identified by unstable Bayes factors across repeated sampling runs. There are improvements to the algorithm that are specifically developed for hierarchical models [@gronau2019simple]. Yet, it remains crucial to assess the stability of Bayes factor estimates.
